\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-30 manuscript/Rmd/draft.Rmd
Rise and shine

Other sports scoring methods

https://www.climbing.com/competition/olympics/a-guide-to-the-olympic-climbing-format/

Soccer points: @haugen2008

Cross country scoring: @hammond2007, @boudreau2018

Decathlon scoring: @westera2006

@@

Brief paragraph on IIA

@@

In this section, we discuss the results of our simulations described in Section \ref{data-and-methods}. For the qualification round, our simulation study shows that a climber is almost guaranteed to make the final round if they win the first event (with a 99.51% chance of advancing) or if they win at least one of the three climbing concentrations (99.48%). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 2. We notice that on average, the minimum score that one should aim for in order to move on to the final round is about 434 (for 8th rank).

@@

We would expect that if a climber with a high combined ranking is removed from the list of competitors, the placement of lower-ranked climbers will shift. In the most trivial case, if we were to remove the 3rd place climber, we would see 4th place climber shift to 3rd, 5th place climber shift to 4th, etc. It is maybe less trivial that when a climber is removed from the list, the placement of their higher-ranking competitors may also shift. In other words, there is a non-zero probability of seeing a change in the placement - particularly of medalists - regardless of the fact that no changes occurred in the performance of the remaining climbers. Real examples are given in section 4.2. In this section, we conduct simulations to analyze the probability of seeing a reordering of medalists when one non-medalist is removed from the list of finalists.

@@

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 2 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau [@kendall1938] as our measure of ordinal association between the quantities. It is evidently clear that there exists a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations. 
<!-- This trend also holds for most of the past competitions. -->

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Men's and Women's Qualifications", fig.height=8, fig.width=6}
mcor <- mq %>% 
  select(Overall = rank, Speed = speed, Bouldering = bould, Lead = lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank()) +
  labs(y = "Men\n")

@@

Another interesting question that we are interested in investigating is "What would happen to the rankings if a single climber is removed?".

This illustrates the idea of IIA

We once again make use of data from the 2018 Youth Olympics for this analysis, but this time we examine the final round of both men's and women's competitions.

Figure 3 shows the plots

A single climber excluded changes things drastically, especially order of medalists.

The cases where someone behind you drops out and your ranking changes. Figure

@@

>>>>

<!-- # https://www.climbing.com/competition/olympics/a-guide-to-the-olympic-climbing-format/ -->

Paragraph on scoring methods in other sports: Soccer points: @haugen2008; Cross country scoring: @hammond2007, @boudreau2018; Decathlon scoring: @westera2006

As a side note, rank product, genes

@@

Paragraph on IIA/Social Choice

Suppose we have 3 people A, B, and C participating in a competition. If A finishes in the first place and C is later disqualified and removed, A should still win. If the original winner (A) loses the modified competition (with C removed), then the Independence of Irrelevant Alternatives has been violated.

Economics, psychology, probability theory. Connect with cross country in previous section

First mentioned by Arrow (1951) https://cowles.yale.edu/sites/default/files/files/pub/mon/m12-all.pdf

Luce (1959) http://www.scholarpedia.org/article/Luce%27s_choice_axiom

and Luce (1977) https://www.imbs.uci.edu/files/personnel/luce/pre1990/1977/Luce_JMP_1977a.pdf

Ray (1973) https://www.jstor.org/stable/1913820

@@

In this section, we discuss the results of our simulations described in Section \ref{data-and-methods}. For the qualification round, our simulation study shows that a climber is almost guaranteed to make the final round if they win the first event or if they win at least one of the three climbing concentrations  (99.51% and 99.48% change of advancing, respectively). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 2. We notice that on average, the minimum score that one should aim for in order to move on to the final round is about 434 (for 8th rank).

@@

Another interesting question that we are interested in investigating is "What would happen to the rankings if a single climber is removed?". This is related to the concept of independence of irrelevant alternatives (IIA), as mentioned in Section \ref{data-and-methods}.

We would expect that if a climber with a high combined ranking is removed from the list of competitors, the placement of lower-ranked climbers will shift. In the most trivial case, if we were to remove the 3rd place climber, we would see 4th place climber shift to 3rd, 5th place climber shift to 4th, etc. It is maybe less trivial that when a climber is removed from the list, the placement of their higher-ranking competitors may also shift. In other words, there is a non-zero probability of seeing a change in the placement - particularly of medalists - regardless of the fact that no changes occurred in the performance of the remaining climbers. Real examples are given in Section \ref{data-analysis}. In this section, we conduct simulations to analyze the probability of seeing a reordering of medalists when one non-medalist is removed from the list of finalists.

<!-- In what percentage of the (6!)^3 orderings would the winner change if one of the 6 competitors was dropped and there were only 5 instead of 6? -->

@@

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 2 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau [@kendall1938] as our measure of ordinal association between the quantities. It is evidently clear that there exists a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations. This trend also holds for most of the past competitions. (add table in the appendix)

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Men's and Women's Qualifications", fig.height=8, fig.width=6.5}
mcor <- mq %>% 
  select(Overall = rank, Speed = speed, Bouldering = bould, Lead = lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank()) +
  labs(y = "Men\n")

@@

Next, we perform analysis on the situations where a climber is dropped from the original standings. We once again make use of data from the 2018 Youth Olympics for this analysis, but this time we examine the final round of both men's and women's competitions.

Figure 3 shows the modified versions of the rankings after each ranked climber is excluded for both gender events. We have clear evidence from this plot that removing a single climber changes the rankings drastically, especially order of medalists. One particular interesting case is where an athlete's position change when someone who originally finished behind them drops out. This situation is illustrated by panel 5 of the women's competition, where the fifth-place climber, Krasovskaia, was excluded; and Meul, whose final rank was fourth in real life, moved up to the second spot and claim the silver medal.

Likewise, dropping a higher-ranked athlete could also result in a major shake-up to the new orderings of players. This type of ranking change is demonstrated by panel 1 and 2 of the men's facet. After dropping the first ranked male contestant, Dohi; Schenk, who previously finished fourth and outside of the medalist group now has the lowest score and become the gold medal winner. A similar situation occurs in the case where the second best competitor (Tanaka) was removed, which results in a jump from fourth place to second place for Pan.

@@

Speed climbers

There is a great dependence on irrelevent party.

### 2021-07-30 manuscript/Rmd/draft.tex
Bed time
Rise and shine

Other sports scoring methods

\url{https://www.climbing.com/competition/olympics/a-guide-to-the-olympic-climbing-format/}

Soccer points: \citet{haugen2008}

Cross country scoring: \citet{hammond2007}, \citet{boudreau2018}

Decathlon scoring: \citet{westera2006}

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Histogram of simulated scores for qualification and final
rounds}
\end{figure}

@@

Brief paragraph on IIA

@@

In this section, we discuss the results of our simulations described in
Section \ref{data-and-methods}. For the qualification round, our
simulation study shows that a climber is almost guaranteed to make the
final round if they win the first event (with a 99.51\% chance of
advancing) or if they win at least one of the three climbing
concentrations (99.48\%). On the other hand, finishing last in the first
event or in any event would certainly hurt an athlete's chance of
finishing in the top 8, as the probabilities of a climber advancing
given they finish last in the first and in any event are 0.1830 and
0.1885, respectively. In addition, the average score for qualification
positions 1 to 8 are displayed in Table 2. We notice that on average,
the minimum score that one should aim for in order to move on to the
final round is about 434 (for 8th rank).

@@

We would expect that if a climber with a high combined ranking is
removed from the list of competitors, the placement of lower-ranked
climbers will shift. In the most trivial case, if we were to remove the
3rd place climber, we would see 4th place climber shift to 3rd, 5th
place climber shift to 4th, etc. It is maybe less trivial that when a
climber is removed from the list, the placement of their higher-ranking
competitors may also shift. In other words, there is a non-zero
probability of seeing a change in the placement - particularly of
medalists - regardless of the fact that no changes occurred in the
performance of the remaining climbers. Real examples are given in
section 4.2. In this section, we conduct simulations to analyze the
probability of seeing a reordering of medalists when one non-medalist is
removed from the list of finalists.

@@

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 2 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau \citep{kendall1938} as our measure of ordinal association
between the quantities. It is evidently clear that there exists a strong
and positive correlation between the ranks of bouldering and lead
climbing, and as a results, the standings of these two events are highly
correlated with the final rankings. On the other hand, the correlation
with the final rank is not as strong for speed climbing. Thus, speed
climbers are facing a huge disadvantage in this scoring system, compared
to those that are specialized in the other two concentrations.

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Men's
and Women's Qualifications}
\end{figure}

@@

Another interesting question that we are interested in investigating is
``What would happen to the rankings if a single climber is removed?''.

This illustrates the idea of IIA

We once again make use of data from the 2018 Youth Olympics for this
analysis, but this time we examine the final round of both men's and
women's competitions.

Figure 3 shows the plots

A single climber excluded changes things drastically, especially order
of medalists.

The cases where someone behind you drops out and your ranking changes.
Figure

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out one
climber. For each gender competition, each panel represents the rank of
the drop-out athlete, with 0 being the original final results. Each case
with a change in rank orderings is highlighted by a black panel border,
and any player with a rank change is represented by a red-filled bar}
\end{figure}

@@

>>>>

Paragraph on scoring methods in other sports: Soccer points:
\citet{haugen2008}; Cross country scoring: \citet{hammond2007},
\citet{boudreau2018}; Decathlon scoring: \citet{westera2006}

As a side note, rank product, genes

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Histogram of simulated scores for qualification and final
rounds}
\end{figure}

@@

Paragraph on IIA/Social Choice

Suppose we have 3 people A, B, and C participating in a competition. If
A finishes in the first place and C is later disqualified and removed, A
should still win. If the original winner (A) loses the modified
competition (with C removed), then the Independence of Irrelevant
Alternatives has been violated.

Economics, psychology, probability theory. Connect with cross country in
previous section

First mentioned by Arrow (1951)
\url{https://cowles.yale.edu/sites/default/files/files/pub/mon/m12-all.pdf}

Luce (1959)
\url{http://www.scholarpedia.org/article/Luce\%27s_choice_axiom}

and Luce (1977)
\url{https://www.imbs.uci.edu/files/personnel/luce/pre1990/1977/Luce_JMP_1977a.pdf}

Ray (1973) \url{https://www.jstor.org/stable/1913820}

@@

In this section, we discuss the results of our simulations described in
Section \ref{data-and-methods}. For the qualification round, our
simulation study shows that a climber is almost guaranteed to make the
final round if they win the first event or if they win at least one of
the three climbing concentrations (99.51\% and 99.48\% change of
advancing, respectively). On the other hand, finishing last in the first
event or in any event would certainly hurt an athlete's chance of
finishing in the top 8, as the probabilities of a climber advancing
given they finish last in the first and in any event are 0.1830 and
0.1885, respectively. In addition, the average score for qualification
positions 1 to 8 are displayed in Table 2. We notice that on average,
the minimum score that one should aim for in order to move on to the
final round is about 434 (for 8th rank).

@@

Another interesting question that we are interested in investigating is
``What would happen to the rankings if a single climber is removed?''.
This is related to the concept of independence of irrelevant
alternatives (IIA), as mentioned in Section \ref{data-and-methods}.

We would expect that if a climber with a high combined ranking is
removed from the list of competitors, the placement of lower-ranked
climbers will shift. In the most trivial case, if we were to remove the
3rd place climber, we would see 4th place climber shift to 3rd, 5th
place climber shift to 4th, etc. It is maybe less trivial that when a
climber is removed from the list, the placement of their higher-ranking
competitors may also shift. In other words, there is a non-zero
probability of seeing a change in the placement - particularly of
medalists - regardless of the fact that no changes occurred in the
performance of the remaining climbers. Real examples are given in
Section \ref{data-analysis}. In this section, we conduct simulations to
analyze the probability of seeing a reordering of medalists when one
non-medalist is removed from the list of finalists.

@@

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 2 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau \citep{kendall1938} as our measure of ordinal association
between the quantities. It is evidently clear that there exists a strong
and positive correlation between the ranks of bouldering and lead
climbing, and as a results, the standings of these two events are highly
correlated with the final rankings. On the other hand, the correlation
with the final rank is not as strong for speed climbing. Thus, speed
climbers are facing a huge disadvantage in this scoring system, compared
to those that are specialized in the other two concentrations. This
trend also holds for most of the past competitions. (add table in the
appendix)

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Men's
and Women's Qualifications}
\end{figure}

@@

Next, we perform analysis on the situations where a climber is dropped
from the original standings. We once again make use of data from the
2018 Youth Olympics for this analysis, but this time we examine the
final round of both men's and women's competitions.

Figure 3 shows the modified versions of the rankings after each ranked
climber is excluded for both gender events. We have clear evidence from
this plot that removing a single climber changes the rankings
drastically, especially order of medalists. One particular interesting
case is where an athlete's position change when someone who originally
finished behind them drops out. This situation is illustrated by panel 5
of the women's competition, where the fifth-place climber, Krasovskaia,
was excluded; and Meul, whose final rank was fourth in real life, moved
up to the second spot and claim the silver medal.

Likewise, dropping a higher-ranked athlete could also result in a major
shake-up to the new orderings of players. This type of ranking change is
demonstrated by panel 1 and 2 of the men's facet. After dropping the
first ranked male contestant, Dohi; Schenk, who previously finished
fourth and outside of the medalist group now has the lowest score and
become the gold medal winner. A similar situation occurs in the case
where the second best competitor (Tanaka) was removed, which results in
a jump from fourth place to second place for Pan.

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out one
climber. For each gender competition, each panel represents the rank of
the drop-out athlete, with 0 being the original final results. Each case
with a change in rank orderings is highlighted by a black panel border,
and any player with a rank change is represented by a red-filled bar}
\end{figure}

@@

Speed climbers

There is a great dependence on irrelevent party.

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

All of the materials related to this manuscript are publicly available
on GitHub at \newline \url{https://github.com/qntkhvn/climbing}.
